Cost- and Deadline-Constrained Provisioning for Scientific Workflow Ensembles in IaaS Clouds

### 2012-04-25 paper/paper.tex
Sections I-V shortened.

% can use linebreaks \\ within to get better formatting as desired
\title{Cost- and Deadline-Constrained Scheduling of Scientific Workflow Ensembles in IaaS Clouds}

@@

\section{Introduction}

Scientific workflows, usually represented as directed acyclic graphs (DAGs), are
an important class of distributed applications that lead to challenging problems
in resource management on grid and utility computing systems. Workflow
applications for large computational problems are often composed of several
inter-related workflows grouped into {\em ensembles}. All the workflows in an
ensemble typically have a similar structure, but they differ in their input
data, number of tasks, and individual task sizes.

The CyberShake~\cite{Callaghan2011} is a good example of an application that
requires scientific workflow ensembles. Generating a hazard map for a region
such as Southern California requires an ensemble of many related workflows to
compute individual hazard curves. The hazard map generated by the CyberShake
group in 2009, for example, required an ensemble of 239 workflows. In astronomy,
for example, a user might wish to use multiple Montage~\cite{Deelman2008}
workflows with different parameters to generate a set of image mosaics that
cover a given area of the sky in different wavelengths, or use an ensemble to generate
the parts of a very large mosaic. The Pegasus team is now working on a Galactic
Plane project that is focused on generating a mosaic of the entire sky. This
ensemble consists of 17 workflows, each of which contains 900 sub-workflows and
computes one ``tile'' of the full mosaic. Another ensemble example is the
Periodograms application~\cite{Vockler2011} that searches for extrasolar planet
by detecting the periodic dips in light intensity. Due to the large scale of the
input data, this application is often split up into multiple batches processed
by different workflow. Additional workflows are created to run the analysis
using different parameters. A recent analysis of data from the Kepler satellite
required three ensembles of 15 workflows.

@@

Infrastructure-as-a-Service (IaaS) clouds offer the ability to provision 
computational resources on-demand according to a pay-per-use model. These systems 
are regarded by the scientific community as a potentially attractive source of 
low-cost, on-demand computing resources~\cite{Ostermann2010, Keahey2009}. In contrast 
to clusters and grids, which typically offer best-effort quality of service, clouds 
give users more flexibility in creating a controlled and managed computing environment.
Clouds provide the ability to adjust resource capacity according to the dynamically changing
computing demands of the application, often called auto-scaling. However, giving users 
more control over the resources also requires the development of new methods for task 
scheduling and resource provisioning. The resource management decisions required 
in cloud scenarios not only have to take into account traditional performance-related 
metrics such as workflow makespan or resource utilization, but must also now consider
budget constraints, since the resources from public (commercial) cloud providers 
usually have monetary costs associated with them~\cite{Durkee2010}.

@@

The main contributions of this paper are:
\begin{itemize}
  \item we define the problem of scheduling prioritized workflows grouped
  into ensembles, under both budget and deadline constraints,
  \item we analyze and develop dynamic (on-line) strategies for task scheduling
  and resource provisioning as well as static algorithms that
  rely on the information about the workflow structure (critical paths and
  workflow levels) and estimates of task runtimes. 
  \item we analyze and develop a hybrid workflow-aware dynamic scheduling
  algorithm, together with admission procedures to predict which workflows from
  the ensemble can be completed,
  \item we evaluate these algorithms using a simulator we have
  developed using CloudSim~\cite{Calheiros2011}, which models the infrastructure
  and the application, taking into account uncertainties in task runtime
  estimations, resource provisioning delays and failures, 
  \item we discuss the performance of the algorithms on a set of
  synthetic benchmark workflow ensembles based on important, real scientific
  applications, in the presence of such of such factors as 
  uncertainties in task runtime estimations, resource provisioning delays and
  failures, on a broad range of different application scenarios and
  varying constraint values.
\end{itemize}

@@

We assume a resource model similar to Amazon's Elastic Compute Cloud (EC2). In this model, virtual machine (VM) instances are requested on-demand. The VMs are billed by the hour, with partial hours being rounded up. Although there are multiple VM types with different amounts of CPU, memory, disk space, and I/O, for this thesis we focus on a single VM type because we assume that for most applications there will typically be only one or two VM types with the best price/performance ratio for the application~\cite{Juve2009}. We also assume that a submitted task has exclusive access to a VM instance and that there is no preemption.

We assume that there is a delay between the time that a new VM instance is requested and when it becomes available to execute tasks. In practice this delay is typically on the order of tens of seconds to a few minutes, and is highly dependent upon the cloud platform and the VM image size~\cite{Nurmi2008b}.

@@

The target applications are ensembles of scientific workflows that can be modeled as Directed Acyclic Graphs (DAGs), where the nodes in the graph represent computational tasks, and the edges represent data- or control-flow dependencies between the tasks.

We assume that there are runtime estimates for each task in the workflow based on either a performance model of the application, or historical data that can be mined. Additionally, we assume that the runtime estimates for individual workflow tasks are not perfect, but may vary based on a uniform distribution of $\pm p\%$.

@@

This work focuses on scheduling and provisioning for workflow ensembles. A workflow ensemble consists of several related workflows that a user wishes to run. Each workflow in an ensemble is given a numeric priority that indicates how important the workflow is to the user. As such, the priorities indicate the utility function of the user. These priorities are absolute in the sense that completing a workflow with a given priority is more valuable (gives the user more utility) than completing all other workflows in the ensemble with lower priorities combined.

The goal of the workflow ensemble scheduling and cloud provisioning problem is to complete as many high-priority workflows as possible given a fixed budget and deadline. Only workflows for which all tasks are finished by the deadline are considered to be complete---partial results are not usable in this model.

@@

DPDS' provisioning procedure is based on resource utilization. DPDS starts with a fixed number of resources calculated based on the available time and budget, and adjusts the number of resources according to how well they are utilized by the application.

Given a budget in dollars $b$, deadline in hours $d$, and the hourly price of a VM in dollars $p$, it is easy to calculate the number of VMs, $N_{VM}$, to provision so that the entire budget is consumed before the deadline:
%
\begin{equation}
\label{eq:static-plan}
N_{VM} = \lceil b / (d * p) \rceil
\end{equation}
%
The DPDS algorithm provisions $N_{VM}$ VMs at the start of the ensemble execution.

As the application runs, the DPDS algorithm periodically computes resource utilization using the percentage of idle VMs over time, and adjusts the number of VMs if the utilization is above or below given thresholds. Because it is assumed that VMs are billed by the hour, DPDS only considers VMs that are approaching their hourly billing cycle when deciding which VMs to terminate. This dynamic provisioning algorithm is shown in Algorithm~\ref{alg:prov}.

\begin{algorithm}[tb]
\caption{Dynamic provisioning algorithm for DPDS}
\label{alg:prov}
{\small
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization threshold; $u_l$: lower utilization
threshold; $v_{max}$: maximum number of VMs
\Procedure{Provision}{}
  \State $V_R\gets$ set of running VMs
    \State $V_C\gets$ set of VMs completing billing cycle
    \State $V_T\gets \emptyset$ \Comment{set of VMs to terminate}
    \State $n_T\gets 0$ \Comment{number of VMs to terminate}
    \If{$ b - c < |V_C| * p\ or\ t > d $ }
      \State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$
      \State $V_T\gets$ select $n_T$ VMs to terminate from $V_C$
      \State \Call{Terminate}{$V_T$} \label{l:terminate1}
    \Else 
    \State $u\gets$ current VM utilization
      \If{$u>u_h$ and $|V_R| < v_{max}*N_{VM}$}
        \State \Call{Start}{$new\ VM$}
      \ElsIf{$u<u_l$}
        \State $V_I\gets$ set of idle VMs
        \State $n_T\gets \lceil|V_I|/2\rceil$ \label{l:nT2}
      \State $V_T\gets$ select $n_T$ VMs to terminate from $V_I$
        \State \Call{Terminate}{$V_T$} \label{l:terminate2}
      \EndIf 
    \EndIf
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

The set of VMs completing their billing cycle is determined by considering both the provisioner interval, and the termination delay of the provider. This guarantees that VMs can be terminated before they start the next billing cycle and prevents the budget from being overrun. The VMs terminated in line \ref{l:terminate1} of Algorithm~\ref{alg:prov} are the ones that would overrun the budget if not terminated in the current provisioning cycle. The VMs terminated in line \ref{l:terminate2} are chosen to increase the resource utilization to the desired threshold. In order to prevent instances from being terminated too quickly, potentially wasting resources that have already been paid for but could be used later, no more than half of the idle resources are terminated during each provisioning cycle. To avoid an uncontrolled increase in the number of instances, which may happen in the case of highly parallel workflows, the provisioner will not start a new VM if the number of running VMs is greater than the product of $N_{VM}$ (from Equation~\ref{eq:static-plan}) and an autoscaling parameter, $v_{max}$. Unless otherwise specified, $v_{max}$ is assumed to be 1.

In order to schedule individual workflow tasks onto available VMs, DPDS uses the dynamic, priority-based scheduling procedure shown in Algorithm~\ref{alg:ds}.

\begin{algorithm}[tb]
\caption{Priority-based scheduling algorithm for DPDS}
\label{alg:ds}
{\small
\begin{algorithmic}[1]
\Procedure{Schedule}{}
    \State $P\gets$ empty priority queue
	\State $IdleVMs\gets$ set of idle VMs
	\For{root task $t$ in all workflows}
    	\State \Call{Insert}{$t,P$} 
    \EndFor
    \While{deadline not reached}
    	\While{$IdleVMs \neq \emptyset\ and\ P \neq \emptyset$}
    		\State $v\gets$ \Call{SelectRandom}{$IdleVMs$}
    		\State $t\gets$ \Call{Pop}{$P$}
    		\State \Call{Submit}{$t,v$}
    	\EndWhile
    	\State Wait for task $t$ to finish on VM $v$
    	\State Update $P$ with ready children of $t$
		\State \Call{Insert}{$v,IdleVMs$}
    \EndWhile
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

Initially, the ready tasks from all workflows in the ensemble are added to a priority queue based on the priority of the workflow to which they belong. If there are idle VMs available, and the priority queue is not empty, the next task from the priority queue is submitted to an arbitrarily chosen idle VM. The process is repeated until there are no idle VMs or the priority queue is empty. The scheduler then waits for a task to finish, adds its ready children to the priority queue, marks the VM as idle, and the entire process repeats until the deadline is reached.

The DPDS algorithm guarantees that tasks from lower priority workflows are always deferred when higher-priority tasks are available, but lower-priority tasks can still occupy idle VMs when higher-priority tasks are not available. However, because there is no preemption, long-running low-priority tasks may delay the execution of higher-priority tasks. In addition, tasks from low priority workflows may be executed even though there is no chance that those workflows will be completed within the current budget and deadline. Figure~\ref{fig:spds-example} shows an example schedule generated using the SPDS algorithm. The figure illustrates how tasks from lower priority workflows backfill idle VMs when tasks from higher priority workflows are not available.

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/spds-gantt}
  \caption[Example schedule generated by the DPDS algorithm.]{Example schedule generated by the DPDS algorithm. Each row is a different VM. Tasks are represented as boxes and are colored by workflow.}
  \label{fig:spds-example}
\end{figure}

@@

\begin{algorithm}[tb]
\caption{Workflow admission algorithm for WA-DPDS}
\label{alg:wa-dpds}
{\small
\begin{algorithmic}[1]
\Require $w$: workflow; $b$: budget; $c$: current cost
\Procedure{Admit}{$w,b,c$}
    \State $r_n\gets b-c$ \Comment{Budget remaining for new VMs}
    \State $r_c\gets $cost committed to VMs that are running
    \State $r_a\gets $cost to complete workflows previously admitted
  \State $r_m\gets 0.1$ \Comment{Safety margin}
  \State $r_b\gets r_n+r_c-r_a-r_m$ \Comment{Budget remaining}
  \State $c_w\gets$ \Call{EstimateCost}{w}
    \If{$c_w<r_b$}
      \State \textbf{return} $TRUE$
    \Else
      \State \textbf{return} $FALSE$
  \EndIf       
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

This admission procedure relies only on the total estimated resource consumption and compares it to the remaining budget. We found that this estimation is useful not only to prevent low-priority workflows from delaying high-priority ones, but also to reject large and costly workflows that would overrun the budget and admit smaller workflows that can efficiently utilize idle resources in ensembles containing workflows of non-uniform sizes. It would also be possible to extend this admission procedure to check other constraints, such as whether the estimated critical path of the new workflow exceeds the time remaining until the deadline.

@@

The previous algorithms are all online algorithms that make provisioning and scheduling decisions at runtime. In comparison, the SPSS algorithm creates a provisioning and scheduling plan before running any workflow tasks. This enables SPSS to start only those workflows that it knows can be completed given the deadline and budget constraints, and eliminates any waste that may be allowed by the dynamic algorithms.

The approach used by SPSS is to plan each workflow in the ensemble in priority order, rejecting any workflows that cannot be completed by the deadline or that cause the cost of the plan to exceed the budget. Once the plan is complete, the VMs are provisioned and tasks are executed according to the schedules given by the plan.

The disadvantage of the static planning approach used by SPSS is that it is sensitive to dynamic changes in the environment and the application that may disrupt its carefully constructed plan. For example, if there are provisioning delays, or if the runtime estimates for the tasks are inaccurate, then workflow execution may diverge from the plan. This issue will be discussed further in Sections~\ref{sec:delays} and \ref{sec:variances}.

Algorithm~\ref{alg:admit} shows how ensembles are planned in SPSS. The algorithm starts with an empty plan containing no VMs and no scheduled tasks. Workflows from the ensemble are considered in priority order. For each workflow, SPSS attempts to build on top of the current plan by provisioning VMs to schedule the tasks of the workflow so that it finishes before the deadline with the least possible cost. If the cost of the new plan is less than the budget, then the new plan is accepted and the workflow is admitted. If not, then the new plan is rejected and the process continues with the next workflow in the ensemble. The idea behind this algorithm is that, if each workflow can be completed by the deadline with the lowest possible cost, then the number of workflows that can be completed within the given budget will be maximized.

\begin{algorithm}[tb]
\caption{Ensemble planning algorithm for SPSS}
\label{alg:admit}
{\small
\begin{algorithmic}[1]
\Require $W$: workflow ensemble; $b$: budget; $d$: deadline
\Ensure Schedule as much of $W$ as possible given $b$ and $d$
\Procedure{PlanEnsemble}{$W,b,d$}
    \State $P\gets \emptyset$ \Comment{Current plan}
    \State $A\gets \emptyset$ \Comment{Set of admitted DAGs}
    \For{$w\ in\ W$}
        \State $P^\prime \gets$\ \Call{PlanWorkflow}{$w,P,d$}
        \If{$Cost(P^\prime) \le b$}
            \State $P\gets\ P^\prime$ \Comment{Accept new plan}
            \State $A \gets A\ +\ w$ \Comment{Admit w}
        \EndIf
    \EndFor
    \State \textbf{return} $P,A$
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

In order to plan a workflow, the SPSS algorithm assigns sub-deadlines to each individual task in the workflow, and then schedules each task so as to minimize the cost of the task while still meeting its assigned sub-deadline. The idea is that if each task can be completed by its deadline in the least expensive manner possible, then the cost of the entire workflow can be minimized without exceeding the ensemble deadline.

\begin{algorithm}[tb]
\caption{Workflow planning algorithm for SPSS}
\label{alg:planworkflow}
{\small
\begin{algorithmic}[1]
\Require $w$: workflow; $P$: current plan; $d$: deadline
\Ensure Create plan for $w$ that minimizes cost and meets deadline $d$
\Procedure{PlanWorkflow}{$w,P,d$}
    \State $P^\prime\gets$ copy of $P$
    \State \Call{DeadlineDistribution}{w,d}
    \For{$t\ in\ w\ sorted\ by\ DL(t)$}
        \State $v \gets$ VM that minimizes cost and start time of t
        \If{$FinishTime(t,v) < DL(t)$}
            \State Schedule(t,v)
        \Else
            \State Provision a new VM v
            \State Schedule(t,v)
        \EndIf
    \EndFor
    \State \textbf{return} $P^\prime$
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

SPSS assigns sub-deadlines to each task based on the slack time of the workflow, which is defined as the amount of extra time that a workflow can extend its critical path and still be completed by the ensemble deadline. For a workflow $w$, the slack time of $w$ is:
%
$$
ST(w) = d - CP(w)
$$
%
where $d$ is the ensemble deadline and $CP(w)$ is the critical path of $w$. We assume that $CP(w) \leq d$, otherwise the workflow cannot be completed by the deadline and must be rejected. For large ensembles we expect the critical path of any individual workflow to be much less than the deadline.

@@

Algorithm~\ref{alg:planworkflow} shows how SPSS creates low-cost plans for each workflow. The \textsc{PlanWorkflow} procedure first calls \textsc{DeadlineDistribution} to assign sub-deadlines to tasks according to the procedure described above. Then, the \textsc{PlanWorkflow} procedure schedules tasks onto VMs, allocating new VMs when necessary. For each task in the workflow, the least expensive slot is chosen to schedule the task so that it can be completed by  its deadline. VMs are allocated in blocks of one billing cycle (one hour) regardless of the size of the task. When computing the cost of scheduling a task on a given VM, the algorithm considers idle slots in blocks that were allocated for previous tasks to be free, while slots in new blocks cost the full price of a billing cycle. For example, if a task has a runtime of 10 minutes, and the price of a block is \$1, then the algorithm will either schedule the task on an existing VM that has an idle slot larger than 10 minutes for a cost of \$0, or it will allocate a new block on an existing VM, or provision a new VM, for a cost of \$1. If the cost of slots on two different VMs is equal, then the slot with the earliest start time is chosen. To prevent too many VMs from being provisioned, the algorithm always prefers to extend the runtime of existing VMs before allocating new VMs. The result of this is that the algorithm will only allocate a new block if there are no idle slots on existing blocks large enough or early enough to complete the task by its deadline, and it will only allocate a new VM if it cannot add a block to an existing VM to complete the task by its deadline.

An example schedule generated by SPSS is shown in Figure.~\ref{fig:spss-example}. This example shows how SPSS tends to start many workflows in parallel, running each workflow over a longer period of time on only a few VMs to minimize cost. In comparison, the dynamic algorithms tend to run one workflow at a time across many VMs in parallel.

>>>>

% can use linebreaks \\ within to get better formatting as desired
\title{Cost- and Deadline-Constrained Provisioning for Scientific Workflow
Ensembles in IaaS Clouds}

@@

\section{Introduction}
\label{sec:intro}

Scientific workflows, usually represented as directed acyclic graphs (DAGs), are
an important class of applications that lead to challenging problems
in resource management on grid and utility computing systems. Workflows
for large computational problems are often composed of several
inter-related workflows grouped into {\em ensembles}. All the workflows in an
ensemble typically have a similar structure, but they differ in their input
data, number of tasks, and individual task sizes.

The CyberShake~\cite{Callaghan2011} is a good example of an application that
requires scientific workflow ensembles. Generating a hazard map for a region
such as Southern California requires an ensemble of many related workflows to
compute individual hazard curves. E.g. the hazard map generated by the
CyberShake group in 2009 required an ensemble of 239 workflows. In astronomy,
for example, a user might wish to use multiple Montage~\cite{Deelman2008}
workflows with different parameters to generate a set of image mosaics that
cover a given area of the sky in different wavelengths, or use an ensemble to generate
the parts of a very large mosaic. The Pegasus team is now working on a Galactic
Plane project that is focused on generating a mosaic of the entire sky. This
ensemble consists of 17 workflows, each of which contains 900 sub-workflows and
computes one ``tile'' of the full mosaic. Another ensemble example is the
Periodograms application~\cite{Vockler2011} that searches for extrasolar planets
by detecting the periodic dips in light intensity. Due to the large scale of the
input data, this application is often split up into multiple batches processed
by different workflows. Additional workflows are created to run the analysis
using different parameters. A recent analysis of data from the Kepler satellite
required three ensembles of 15 workflows.

@@

Infrastructure-as-a-Service (IaaS) clouds offer the ability to provision 
computational resources on-demand according to a pay-per-use model. These systems 
are regarded by the scientific community as a potentially attractive source of 
low-cost computing resources~\cite{Ostermann2010, Keahey2009}. In contrast 
to clusters and grids, which typically offer best-effort quality of service, clouds 
give more flexibility in creating a controlled and managed computing environment.
Clouds provide the ability to adjust resource capacity according to the changing
demands of the application, often called auto-scaling. However, giving users 
more control also requires the development of new methods for task 
scheduling and resource provisioning. The resource management decisions required 
in cloud scenarios not only have to take into account traditional performance-related 
metrics such as workflow makespan or resource utilization, but must also now consider
budget constraints, since the resources from public (commercial) clouds
usually have monetary costs associated with them~\cite{Durkee2010}.

@@

The main contributions of this paper are:
\begin{itemize}
  \item we define the problem of scheduling prioritized workflows grouped
  into ensembles, under both budget and deadline constraints,
  \item we analyze and develop dynamic (on-line) strategies for task scheduling
  and resource provisioning as well as static algorithms that
  rely on the information about the workflow structure (critical paths and
  workflow levels) and estimates of task runtimes. 
  \item we analyze and develop a hybrid workflow-aware dynamic scheduling
  algorithm, together with admission procedures to predict which workflows from
  the ensemble can be completed,
  \item we evaluate these algorithms using our simulator based on
  CloudSim~\cite{Calheiros2011}, which models the infrastructure and the application, 
  taking into account uncertainties in task runtime
  estimations, provisioning delays and failures, 
  \item we discuss the performance of the algorithms on a set of
  synthetic workflow ensembles based on important, real scientific
  applications, on a broad range of different application scenarios and
  varying constraint values.
\end{itemize}

@@

We assume a resource model similar to Amazon's Elastic Compute Cloud (EC2),
where virtual machine (VM) instances are requested on-demand and are billed by
the hour, with partial hours being rounded up. Although there may be
heterogeneous VM types with different amounts of CPU, memory, disk space, and
I/O, for this paper we focus on a single VM type because we assume that for most
applications there will typically be only one or two VM types with the best
price/performance ratio for the application~\cite{Juve2009}. We also assume that
a submitted task has exclusive access to a VM instance and that there is no
preemption. We assume that there is a delay between the time that a new VM
instance is requested and when it becomes available to execute tasks. In
practice this delay is typically on the order of tens of seconds to a few
minutes, and is highly dependent upon the cloud platform and the VM image
size~\cite{Nurmi2008b}.

@@

The target applications are ensembles of scientific workflows that can be
modeled as Directed Acyclic Graphs (DAGs), where the nodes in the graph
represent computational tasks, and the edges represent data- or control-flow
dependencies between the tasks. We assume that there are runtime estimates for
each task in the workflow based on either a performance model of the
application, or historical data that can be mined. Additionally, we assume that
the runtime estimates for individual workflow tasks are not perfect, but may
vary based on a uniform distribution of $\pm p\%$.

@@

A workflow ensemble consists of several related workflows that a user wishes to
run. Each workflow in an ensemble is given a numeric priority that indicates how
important the workflow is to the user. As such, the priorities indicate the
utility function of the user. These priorities are absolute in the sense that
completing a workflow with a given priority is more valuable (gives the user
more utility) than completing all other workflows in the ensemble with lower
priorities combined. The goal of the workflow ensemble scheduling and cloud
provisioning problem is to complete as many high-priority workflows as possible
given a fixed budget and deadline. Only workflows for which all tasks are
finished by the deadline are considered to be complete---partial results are not
usable in this model.

@@

DPDS' provisioning procedure is based on resource utilization. DPDS starts with
a fixed number of resources calculated based on the available time and budget,
and adjusts the number of resources according to how well they are utilized by
the application.
Given a budget in dollars $b$, deadline in hours $d$, and the hourly price of a
VM in dollars $p$, it is easy to calculate the number of VMs, $N_{VM}$, to
provision so that the entire budget is consumed before the deadline:
\begin{equation}
\label{eq:static-plan}
N_{VM} = \lceil b / (d * p) \rceil
\end{equation}
%
The DPDS algorithm provisions $N_{VM}$ VMs at the start of the ensemble execution.

The DPDS algorithm periodically computes resource
utilization using the percentage of idle VMs over time, and adjusts the number
of VMs if the utilization is above or below given thresholds. Because it is
assumed that VMs are billed by the hour, DPDS only considers VMs that are
approaching their hourly billing cycle when deciding which VMs to terminate.
This dynamic provisioning algorithm is shown in Algorithm~\ref{alg:prov}.

\begin{algorithm}[tb]
\caption{Dynamic provisioning algorithm for DPDS}
\label{alg:prov}
{\footnotesize
\begin{algorithmic}[1]
\Require $c$: consumed budget; $b$: total budget; $d$: deadline; $p$: price;
$t$: current time; $u_h$: upper utilization threshold; $u_l$: lower utilization
threshold; $v_{max}$: maximum number of VMs
\Procedure{Provision}{}
  \State $V_R\gets$ set of running VMs
    \State $V_C\gets$ set of VMs completing billing cycle
    \State $V_T\gets \emptyset$ \Comment{set of VMs to terminate}
    \State $n_T\gets 0$ \Comment{number of VMs to terminate}
    \If{$ b - c < |V_C| * p\ or\ t > d $ }
      \State $n_T\gets |V_R| - \lfloor(b-c)/p\rfloor$
      \State $V_T\gets$ select $n_T$ VMs to terminate from $V_C$
      \State \Call{Terminate}{$V_T$} \label{l:terminate1}
    \Else 
    \State $u\gets$ current VM utilization
      \If{$u>u_h$ and $|V_R| < v_{max}*N_{VM}$}
        \State \Call{Start}{$new\ VM$}
      \ElsIf{$u<u_l$}
        \State $V_I\gets$ set of idle VMs
        \State $n_T\gets \lceil|V_I|/2\rceil$ \label{l:nT2}
      \State $V_T\gets$ select $n_T$ VMs to terminate from $V_I$
        \State \Call{Terminate}{$V_T$} \label{l:terminate2}
      \EndIf 
    \EndIf
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

The set of VMs completing their billing cycle is determined by  both
the provisioner interval, and the termination delay of the provider. This
guarantees that VMs can be terminated before they start the next billing cycle
and prevents the budget from being overrun. The VMs terminated in line
\ref{l:terminate1} of Algorithm~\ref{alg:prov} are the ones that would overrun
the budget if not terminated in the current provisioning cycle. The VMs
terminated in line \ref{l:terminate2} are chosen to increase the resource
utilization to the desired threshold. In order to prevent instances that have
already been paid for from being terminated too quickly, no more than half of
the idle resources are terminated during each provisioning cycle. To avoid an
uncontrolled increase in the number of instances, which may happen in the case
of highly parallel workflows, the provisioner will not start a new VM if the
number of running VMs is greater than the product of $N_{VM}$ (from
Equation~\ref{eq:static-plan}) and an autoscaling parameter, $v_{max}$. Unless
otherwise specified, $v_{max}$ is assumed to be 1.

In order to schedule individual workflow tasks onto available VMs, DPDS uses the
dynamic, priority-based scheduling procedure shown in Algorithm~\ref{alg:ds}.
Initially, the ready tasks from all workflows in the ensemble are added to a
priority queue based on the priority of the workflow to which they belong. If
there are idle VMs available, and the priority queue is not empty, the next task
from the priority queue is submitted to an arbitrarily chosen idle VM. The
process is repeated until there are no idle VMs or the priority queue is empty.
The scheduler then waits for a task to finish, adds its ready children to the
priority queue, marks the VM as idle, and the entire process repeats until the
deadline is reached.

\begin{algorithm}[tb]
\caption{Priority-based scheduling algorithm for DPDS}
\label{alg:ds}
{\footnotesize
\begin{algorithmic}[1]
\Procedure{Schedule}{}
    \State $P\gets$ empty priority queue
	\State $IdleVMs\gets$ set of idle VMs
	\For{root task $t$ in all workflows}
    	\State \Call{Insert}{$t,P$} 
    \EndFor
    \While{deadline not reached}
    	\While{$IdleVMs \neq \emptyset\ and\ P \neq \emptyset$}
    		\State $v\gets$ \Call{SelectRandom}{$IdleVMs$}
    		\State $t\gets$ \Call{Pop}{$P$}
    		\State \Call{Submit}{$t,v$}
    	\EndWhile
    	\State Wait for task $t$ to finish on VM $v$
    	\State Update $P$ with ready children of $t$
		\State \Call{Insert}{$v,IdleVMs$}
    \EndWhile
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

The DPDS algorithm guarantees that tasks from lower priority workflows are
always deferred when higher-priority tasks are available, but lower-priority
tasks can still occupy idle VMs when higher-priority tasks are not available.
However, because there is no preemption, long-running low-priority tasks may
delay the execution of higher-priority tasks. In addition, tasks from low
priority workflows may be executed even though there is no chance that those
workflows will be completed within the current budget and deadline.
Figure~\ref{fig:algorithm-example}a shows an example schedule generated using
the DPDS algorithm. The figure illustrates how tasks from lower priority workflows
backfill idle VMs when tasks from higher priority workflows are not available.

@@

\begin{algorithm}[tb]
\caption{Workflow admission algorithm for WA-DPDS}
\label{alg:wa-dpds}
{\footnotesize
\begin{algorithmic}[1]
\Require $w$: workflow; $b$: budget; $c$: current cost
\Procedure{Admit}{$w,b,c$}
    \State $r_n\gets b-c$ \Comment{Budget remaining for new VMs}
    \State $r_c\gets $cost committed to VMs that are running
    \State $r_a\gets $cost to complete workflows previously admitted
  \State $r_m\gets 0.1$ \Comment{Safety margin}
  \State $r_b\gets r_n+r_c-r_a-r_m$ \Comment{Budget remaining}
  \State $c_w\gets$ \Call{EstimateCost}{w}
    \If{$c_w<r_b$}  \textbf{return} $TRUE$
    \Else ~  \textbf{return} $FALSE$
  \EndIf       
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

This admission procedure relies only on the total estimated resource consumption
and compares it to the remaining budget. We found that this estimation is useful
not only to prevent low-priority workflows from delaying high-priority ones, but
also to reject large and costly workflows that would overrun the budget and
admit smaller workflows that can efficiently utilize idle resources.
% in
%ensembles containing workflows of non-uniform sizes. It would also be possible
%to extend this admission procedure to check other constraints, such as whether
%the estimated critical path of the new workflow exceeds the time remaining
%until the deadline.

@@

The previous dynamic (online) algorithms make provisioning and scheduling
decisions at runtime. In comparison, the SPSS algorithm creates a provisioning
and scheduling plan before running any workflow tasks. This enables SPSS to
start only those workflows that it knows can be completed given the deadline and
budget constraints, and eliminates any waste that may be allowed by the dynamic
algorithms. However, the disadvantage of SPSS is that it is sensitive to dynamic
changes in the environment (see Sections~\ref{sec:delays} and
\ref{sec:variances}).

The approach used by SPSS is to plan each workflow in the ensemble in priority
order, rejecting any workflows that would exceed the dedline or budget. Once the
plan is complete, the VMs are provisioned and tasks are executed according to
the schedules given by the plan.

%The disadvantage of the static planning approach used by SPSS is that it is
%sensitive to dynamic changes in the environment and the application that may
%disrupt its carefully constructed plan. For example, if there are provisioning
%delays, or if the runtime estimates for the tasks are inaccurate, then workflow
%execution may diverge from the plan. This issue will be discussed further in
%Sections~\ref{sec:delays} and \ref{sec:variances}.

Algorithm~\ref{alg:admit} shows how ensembles are planned in SPSS. Workflows
from the ensemble are considered in priority order. For each workflow, SPSS
attempts to build on top of the current plan by provisioning VMs to schedule the
tasks of the workflow so that it finishes before the deadline with the least
possible cost. If the cost of the new plan is less than the budget, then the new
plan is accepted and the workflow is admitted. If not, then the new plan is
rejected and the process continues with the next workflow in the ensemble. The
idea is that, if each workflow can be completed by the
deadline with the lowest possible cost, then the number of workflows that can be
completed within the given budget will be maximized.

To plan a workflow, the SPSS algorithm assigns sub-deadlines to each individual
task in the workflow, and then schedules each task so as to minimize the cost of
the task while still meeting its assigned sub-deadline. If each task can be
completed by its deadline in the least expensive way, then the cost of the
entire workflow can be minimized without exceeding the deadline. SPSS assigns
sub-deadlines to each task based on the slack time of the workflow, which is
defined as the amount of extra time that a workflow can extend its critical path
and still be completed by the ensemble deadline. For a workflow $w$, the slack
time of $w$ is:
$$ ST(w) = d - CP(w) $$
where $d$ is the deadline and $CP(w)$ is the critical path of $w$. We
assume that $CP(w) \leq d$, otherwise the workflow cannot be completed by the
deadline and must be rejected. 
%For large ensembles we expect the critical path
%of any individual workflow to be much less than the deadline.

\begin{algorithm}[tb]
\caption{Ensemble planning algorithm for SPSS}
\label{alg:admit}
{\footnotesize
\begin{algorithmic}[1]
\Require $W$: workflow ensemble; $b$: budget; $d$: deadline
\Ensure Schedule as much of $W$ as possible given $b$ and $d$
\Procedure{PlanEnsemble}{$W,b,d$}
    \State $P\gets \emptyset$ \Comment{Current plan}
    \State $A\gets \emptyset$ \Comment{Set of admitted DAGs}
    \For{$w\ in\ W$}
        \State $P^\prime \gets$\ \Call{PlanWorkflow}{$w,P,d$}
        \If{$Cost(P^\prime) \le b$}
            \State $P\gets\ P^\prime$ \Comment{Accept new plan}
            \State $A \gets A\ +\ w$ \Comment{Admit w}
        \EndIf
    \EndFor
    \State \textbf{return} $P,A$
\EndProcedure
\end{algorithmic}
}
\end{algorithm}

@@

\begin{algorithm}[tb]
\caption{Workflow planning algorithm for SPSS}
\label{alg:planworkflow}
{\footnotesize
\begin{algorithmic}[1]
\Require $w$: workflow; $P$: current plan; $d$: deadline
\Ensure Create plan for $w$ that minimizes cost and meets deadline $d$
\Procedure{PlanWorkflow}{$w,P,d$}
    \State $P^\prime\gets$ copy of $P$
    \State \Call{DeadlineDistribution}{w,d}
    \For{$t\ in\ w\ sorted\ by\ DL(t)$}
        \State $v \gets$ VM that minimizes cost and start time of t
        \If{$FinishTime(t,v) < DL(t)$}
            \State Schedule(t,v)
        \Else
            \State Provision a new VM v
            \State Schedule(t,v)
        \EndIf
    \EndFor
    \State \textbf{return} $P^\prime$
\EndProcedure
\end{algorithmic}
}
\end{algorithm}
